---
title: "Multimorbidity Patterns"
author: "for Behavioral Health Patients in the Lakeshore Region"
output: html_document
---

## Context

The number of chronic conditions that a person has: 

* affects behavioral health costs [^1]
* increases mortality [^2]
* increases inpatient and long-term care admissions 

How can we better integrate and coordinate behavioral and medical health services to address these complex issues?


```{r}
summary(cars)
```

## Application

### Care Coordination
Large itemsets (>4) may be less actionable for cross-disciplinary efforts, especially when these diagnoses fall into different medical specialties, due to the difficulty in coordinating efforts across multiple parties for increasingly small numbers of patients.  That said, the larger itemsets do paint a picture which is helpful for care coordination.

### Analytics Platform
Display

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```


---
## Notes
[^1]: [Freeman, E., McGuire, C. A., Thomas, J. W., & Thayer, D. A. (2014). Factors affecting costs in Medicaid populations with behavioral health disorders. Medical care, 52, S60-S66.](http://journals.lww.com/lww-medicalcare/Fulltext/2014/03001/Factors_Affecting_Costs_in_Medicaid_Populations.12.aspx)
[^2]: [Zeng, C., Ellis, J. L., Steiner, J. F., Shoup, J. A., McQuillan, D. B., & Bayliss, E. A. (2014). Assessment of morbidity over time in predicting health outcomes. Medical care, 52, S52-S59.](http://journals.lww.com/lww-medicalcare/Fulltext/2014/03001/Assessment_of_Morbidity_Over_Time_in_Predicting.11.aspx)