---
title: "Multimorbidity Patterns"
author: "for Behavioral Health Patients in the Lakeshore Region"
output: html_document
---

```{r, echo=FALSE,message=FALSE,warning=FALSE}

```


## Context

Anyone who has provided behavioral health services to an individual understands that diagnoses do not limit their impact to one part of a person's life, but influence multiple areas of a person's day-to-day existence.  Multiple diagnoses, especially chronic conditions, have a cumulative impact across a person's life.  There is a systemic and economic effect as well.  Research indicates that the number of chronic conditions a person has: 

* affects behavioral health costs [^1]
* increases mortality [^2]
* increases inpatient and long-term care admissions 

## Questions

What are the most prevalent multimorbidity patterns in the specialty behavioral health system for this region?  

Which of these multimorbidity patterns accounts for the greatest proportion of spending in the region?  Of service use?

Of high-cost/use multimorbidity patterns, which have a strong evidence base for effective treatments?

Of multimorbidity patterns with indications for effective treatment, what proportion of patients with these patterns are receiving services from providers which may indicate the provision of this treatment? (e.g. by provider type or service code)

How can we better integrate and coordinate behavioral and medical health services to address these complex issues?

Of individuals with multimorbidity patterns with high service use, how many are receiving care management?



## Data
```{r}
summary(cars)
```

## Methodology

## Application
The question of "what items commonly appear together?" is a prevalent one.  It has been asked by many retailers in regards to goods that wind up in customers' "carts" by the time they check out and pay---i.e., the items most often purchased together in a transaction.  The apriori and eclat algorithms were developed for this sort of transactional analysis.  Algorithms written in C by Christian Borgelt have been ported to R, where they are accessed via the arules package.\footnote{The arules package Version 1.1-6 is authored by Michael Hahsler, Christian Buchta, Bettina Gruen, Kurt Hornik, and Christian Borgelt.  Information about this package is available at the url
http://r-forge.r-project.org/projects/arules/
The package is available for download and inclusion in one's R installation from various CRAN mirror sites.}

In our analysis, we took a "transaction" to correspond to available data for a single patient.  More specifically, every patient generated one transaction, and the items in that transaction were the conditions reported for that patient.  We, thus, begin with transactional data which might be as depicted in the table below, where each row represents a single patient/transaction.

Insert a table here.

A long list of transactions was generated from the raw data, and used as input to the eclat() function.  The eclat algorithm determined the which comorbidity patterns were most prevalent.  In calling eclat(), it was necessary to specify several parameters, including
 * support: just what portion of participants
   should have a comorbidity condition before
 * minlen: the minimum number of conditions an
   itemset deemed "frequent" should contain.


### Care Coordination
Large itemsets (>4) may be less actionable for cross-disciplinary efforts, especially when these diagnoses fall into different medical specialties, due to the difficulty in coordinating efforts across multiple parties for increasingly small numbers of patients.  That said, the larger itemsets do paint a picture which is helpful for care coordination.

### Analytics Platform
Display


---
## Notes
[^1]: [Freeman, E., McGuire, C. A., Thomas, J. W., & Thayer, D. A. (2014). Factors affecting costs in Medicaid populations with behavioral health disorders. Medical care, 52, S60-S66.](http://journals.lww.com/lww-medicalcare/Fulltext/2014/03001/Factors_Affecting_Costs_in_Medicaid_Populations.12.aspx)
[^2]: [Zeng, C., Ellis, J. L., Steiner, J. F., Shoup, J. A., McQuillan, D. B., & Bayliss, E. A. (2014). Assessment of morbidity over time in predicting health outcomes. Medical care, 52, S52-S59.](http://journals.lww.com/lww-medicalcare/Fulltext/2014/03001/Assessment_of_Morbidity_Over_Time_in_Predicting.11.aspx)