---
title: "Multimorbidity Patterns"
author: "for Behavioral Health Patients in the Lakeshore Region"
output:
  html_document: 
    toc: true
    toc_depth: 1
    theme: cosmo 
    highlight: tango
---

```{r, echo=FALSE,message=FALSE,warning=FALSE}
source("loadTransactions.R")
source("processDat.R")
```


## Context

Anyone who has provided behavioral health services to an individual understands that diagnoses do not limit their impact to one part of a person's life, but influence multiple areas of a person's day-to-day existence.  Multiple diagnoses, especially chronic conditions, have a cumulative impact across a person's life.  There is a systemic and economic effect as well.[^1]  Research indicates that the number of chronic conditions which a person has: 

* affects behavioral health costs [^2]
* increases mortality [^3]
* increases inpatient and long-term care admissions 

A recent analysis of super-utilizer group characteristics found that 82% of super-utilizers had multiple comorbid chronic conditions.[^4] Knowing the number of conditions that a person has, however, does not provide enough insight to help population managers and service providers target clinical strategies for these high-priority populations.  In order to begin to develop proficiency in managing populations with complex chronic conditions, it will be critical to become aware of the patterns in which these conditions occur.  In other words, we can't treat "3 chronic conditions" but we can develop a strategy to address "cardiac dysrythmias, depression and hypertension".

This analysis is intended as an initial step to help the PIHP to begin to target and prioritize specific opportunities for the development of integrated models for beneficiaries with serious mental illness, substance use disorders, developmental disabilities and chronic physical conditions.

## Questions

What are the most prevalent multimorbidity patterns in the specialty behavioral health system for this region?  

How many of the frequent itemsets 

Which of these multimorbidity patterns accounts for the greatest proportion of spending in the region?  Of service use?

How many people have multiple conditions with similar symptomology and thus are potential candidates for differential diagnosis and revised tx planning?

Of high-cost/use multimorbidity patterns, which have a strong evidence base for effective treatments?

Of multimorbidity patterns with indications for effective treatment, what proportion of patients with these patterns are receiving services from providers which may indicate the provision of this treatment? (e.g. by provider type or service code)

How can we better integrate and coordinate behavioral and medical health services to address these complex issues?

Of individuals with multimorbidity patterns with high service use, how many are receiving care management?



## Data
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(magrittr)
library(sunburstR)

# Clean and format data
  tst <- freqItemsets
  tst$items <- gsub("{|}| ","",tst$items, perl=T) # remove brackets
  spl <- strsplit(as.character(tst$items), split = ",")
  n <- max(sapply(spl, length))
  l <- lapply(spl, function(X) c(X, rep(NA, n - length(X))))
  d <- data.frame(t(do.call(cbind, l)))
  d$X1 <- stringr::str_sub(d$X1, start = 1, end = 13)
  d$X2 <- stringr::str_sub(d$X2, start = 1, end = 13)
  d$X3 <- stringr::str_sub(d$X3, start = 1, end = 13)
  d$X4 <- stringr::str_sub(d$X4, start = 1, end = 13)
  # Remove NAs from 3rd dx position before pasting
  d$paths <- ifelse(!is.na(d$X3), 
                    paste(d$X1,d$X2,d$X3,sep = "-"), 
                    paste(d$X1,d$X2,sep = "-"))
  # Remove NAs from 4th dx position before pasting
  d$paths <- ifelse(!is.na(d$X4), 
                    paste(d$paths,d$X4,sep = "-"), 
                    d$paths)
  tst <- cbind(freqItemsets,d)
  tst$paths <- tolower(tst$paths)
  tst$patients <- tst$support * pCount
  tst$patients <- as.integer(tst$patients)
  tst$paths <- as.factor(tst$paths)
  tst %<>% select(paths, counts = patients)

sunburst(tst)

```

## Methodology

## Application
The question of "what items commonly appear together?" is a prevalent one.  It has been asked by many retailers in regards to goods that wind up in customers' "carts" by the time they check out and pay---i.e., the items most often purchased together in a transaction.  The apriori and eclat algorithms were developed for this sort of transactional analysis.  Algorithms written in C by Christian Borgelt have been ported to R, where they are accessed via the arules package.\footnote{The arules package Version 1.1-6 is authored by Michael Hahsler, Christian Buchta, Bettina Gruen, Kurt Hornik, and Christian Borgelt.  Information about this package is available at the url
http://r-forge.r-project.org/projects/arules/
The package is available for download and inclusion in one's R installation from various CRAN mirror sites.}

In our analysis, we took a "transaction" to correspond to available data for a single patient.  More specifically, every patient generated one transaction, and the items in that transaction were the conditions reported for that patient.  We, thus, begin with transactional data which might be as depicted in the table below, where each row represents a single patient/transaction.

Insert a table here.

A long list of transactions was generated from the raw data, and used as input to the eclat() function.  The eclat algorithm determined the which comorbidity patterns were most prevalent.  In calling eclat(), it was necessary to specify several parameters, including
 * support: just what portion of participants
   should have a comorbidity condition before
 * minlen: the minimum number of conditions an
   itemset deemed "frequent" should contain.


```{r, echo=FALSE, warning=FALSE, message=FALSE}
tst <- freqItemsets
tst$items <- gsub("{|}","",tst$items, perl=T) # remove brackets
spl <- strsplit(as.character(tst$items), split = ",")
n <- max(sapply(spl, length))
l <- lapply(spl, function(X) c(X, rep(NA, n - length(X))))
d <- data.frame(t(do.call(cbind, l)))
d$search1 <- tolower(d$X1)
d$search2 <- tolower(d$X2)
d$search3 <- tolower(d$X3)
d$search4 <- tolower(d$X4)

# Use Systematic Review query logic
# http://www.nlm.nih.gov/bsd/pubmed_subsets/sysreviews_strategy.html

query <- paste0("(",d$search1," AND ",d$search2,") ",
               ifelse(is.na(d$search3),
                      yes = "",
                      no = paste("OR (",d$search1," AND ",d$search3,") ")),
               ifelse(is.na(d$search3),
                      yes = "",
                      no = paste("OR (",d$search2," AND ",d$search3,") ")),
               "AND systematic[sb] ") # ,"AND comorbidity[mh]"

d <- cbind(d,query)

d$link <- paste0("http://www.ncbi.nlm.nih.gov/pubmed/?term=",
                 gsub(pattern = " ", replacement = "+", x = d$query))

d <- cbind(freqItemsets,d)

library(DT)

d %>% 
  mutate(ref = paste0("<a href=",link,">",items,"</a>"),
         pct = round(support * 100, digits = 1)) %>%
  select(ref,pct) %>%
  datatable(rownames = FALSE,
                colnames = c('Multimorbidity Pattern', '% of Patients'),
                escape = FALSE)
```


### Care Management
There are a number of reasons why care management interventions have not consistently been shown to decrease healthcare costs of high utilizers[^5]. One reason is that these patients are clinically heterogeneous, and their conditions are not consistently responsive to treatment. 

### Care Coordination
Large itemsets (>4) may be less actionable for cross-disciplinary efforts, especially when these diagnoses fall into different medical specialties, due to the difficulty in coordinating efforts across multiple parties for increasingly small numbers of patients.  That said, the larger itemsets do paint a picture which is helpful for defining common use cases for coordinating communication between multiple specialists.

### Analytics Platform
Display


---
## Notes
[^1]: This report is indebted to research conducted by Johns Hopkins and published in [C. Boyd, B. Leff, C. Weiss, J. Wolff, R. Clark, and T. Richards. Full Report: Clarifying Multimorbidity to Improve Targeting and Delivery of Clinical Services for Medicaid Populations. Center for Health Care Strategies, Inc. December 2010.](http://www.chcs.org/media/Clarifying_Multimorbidity_for_Medicaid_report-FINAL.pdf)
[^2]: [Freeman, E., McGuire, C. A., Thomas, J. W., & Thayer, D. A. (2014). Factors affecting costs in Medicaid populations with behavioral health disorders. Medical care, 52, S60-S66.](http://journals.lww.com/lww-medicalcare/Fulltext/2014/03001/Factors_Affecting_Costs_in_Medicaid_Populations.12.aspx)
[^3]: [Zeng, C., Ellis, J. L., Steiner, J. F., Shoup, J. A., McQuillan, D. B., & Bayliss, E. A. (2014). Assessment of morbidity over time in predicting health outcomes. Medical care, 52, S52-S59.](http://journals.lww.com/lww-medicalcare/Fulltext/2014/03001/Assessment_of_Morbidity_Over_Time_in_Predicting.11.aspx)
[^4]: [Johnson, T. L., Rinehart, D. J., Durfee, J., Brewer, D., Batal, H., Blum, J., ... & Gabow, P. (2015). For Many Patients Who Use Large Amounts Of Health Care Services, The Need Is Intense Yet Temporary. Health Affairs, 34(8), 1312-1319.](http://content.healthaffairs.org/content/34/8/1312.full)
[^5]: [Williams, Brent C. (2015). Limited Effects of Care Management for High Utilizers on Total Healthcare Costs. AJMC. Published Online: May 07, 2015](http://www.ajmc.com/journals/issue/2015/2015-vol21-n4/limited-effects-of-care-management-for-high-utilizers-on-total-healthcare-costs)
