---
title: "Multimorbidity Patterns"
author: "for Behavioral Health Patients in the Lakeshore Region"
output: html_document
---

```{r, echo=FALSE,message=FALSE,warning=FALSE}

```


## Context

Anyone who has provided behavioral health services to an individual understands that diagnoses do not limit their impact to one part of a person's life, but influence multiple areas of a person's day-to-day existence.  Multiple diagnoses, especially chronic conditions, have a cumulative impact across a person's life.  There is a systemic and economic effect as well.[^1]  Research indicates that the number of chronic conditions a person has: 

* affects behavioral health costs [^2]
* increases mortality [^3]
* increases inpatient and long-term care admissions 

Knowing the number of conditions that a person has, however, does not provide enough insight to help population managers and service providers target clinical strategies for these high-priority populations.  In order to begin to develop proficiency in managing populations with complex chronic conditions, it will be critical to become aware of the patterns in which these conditions occur.  In other words, we can't treat "3 chronic conditions" but we can develop a strategy to address "cardiac dysrythmias, depression and hypertension".

This analysis is intended as an initial step to help the PIHP to begin to target and prioritize specific opportunities for the development of integrated models for beneficiaries with serious mental illness, substance use disorders, developmental disabilities and chronic physical conditions.

## Questions

What are the most prevalent multimorbidity patterns in the specialty behavioral health system for this region?  

How many of the frequent itemsets 

Which of these multimorbidity patterns accounts for the greatest proportion of spending in the region?  Of service use?

Of high-cost/use multimorbidity patterns, which have a strong evidence base for effective treatments?

Of multimorbidity patterns with indications for effective treatment, what proportion of patients with these patterns are receiving services from providers which may indicate the provision of this treatment? (e.g. by provider type or service code)

How can we better integrate and coordinate behavioral and medical health services to address these complex issues?

Of individuals with multimorbidity patterns with high service use, how many are receiving care management?



## Data
```{r}
summary(cars)
```

## Application

### Care Coordination
Large itemsets (>4) may be less actionable for cross-disciplinary efforts, especially when these diagnoses fall into different medical specialties, due to the difficulty in coordinating efforts across multiple parties for increasingly small numbers of patients.  That said, the larger itemsets do paint a picture which is helpful for care coordination.

### Analytics Platform
Display


---
## Notes
[^1]: This report is indebted to research conducted by Johns Hopkins and published in [C. Boyd, B. Leff, C. Weiss, J. Wolff, R. Clark, and T. Richards. Full Report: Clarifying Multimorbidity to Improve Targeting and Delivery of Clinical Services for Medicaid Populations. Center for Health Care Strategies, Inc. December 2010.](http://www.chcs.org/media/Clarifying_Multimorbidity_for_Medicaid_report-FINAL.pdf)
[^1]: [Freeman, E., McGuire, C. A., Thomas, J. W., & Thayer, D. A. (2014). Factors affecting costs in Medicaid populations with behavioral health disorders. Medical care, 52, S60-S66.](http://journals.lww.com/lww-medicalcare/Fulltext/2014/03001/Factors_Affecting_Costs_in_Medicaid_Populations.12.aspx)
[^2]: [Zeng, C., Ellis, J. L., Steiner, J. F., Shoup, J. A., McQuillan, D. B., & Bayliss, E. A. (2014). Assessment of morbidity over time in predicting health outcomes. Medical care, 52, S52-S59.](http://journals.lww.com/lww-medicalcare/Fulltext/2014/03001/Assessment_of_Morbidity_Over_Time_in_Predicting.11.aspx)